```{r, include=FALSE}
library(tidyverse)
```

# Logit Model for Qualitative Predictors

```{r, message=FALSE}
aids <- read_table("data/AIDS.dat")
```

<!-- change to format = latex before compiling -->

```{r, echo=FALSE, results='asis'}
knitr::kable(aids %>% group_by(race), format = "html", row.names = FALSE, booktabs = TRUE,
             col.names = c("Race", "AZT Use", "Yes", "No")) %>% 
  kableExtra::kable_styling(latex_options = "striped") %>% 
  kableExtra::add_header_above(header = c(" ", " ", "Symptoms" = 2)) %>% 
  kableExtra::collapse_rows(columns = 1)
```

$$Y = \text{symptoms} = \begin{cases} \text{yes} \\ \text{no} \end{cases}$$

Here, our main interest is whether the usage of AZT(`azt`) would affect this binary response. Denote that the `azt` is also categorical predictor.

$$X = \text{AZT} = \begin{cases} \text{yes} \\ \text{no} \end{cases}$$

There is another factor `race` that has possibility to be covariate.

$$Z = \text{Race} = \begin{cases} \text{White} \\ \text{Black} \end{cases}$$

Based on the interest, we need to control the effect of this covariate.

## ANOVA-Type Representation of Factors

### One-way ANOVA representation

First consider a signle factor case, with $I$ categories(here, $I = 2$).

```{r}
aids %>% 
  select(-race)
```

For each row $i$ of the table, denote

$$
\begin{cases}
n_i = \text{yes} + \text{no} \\
y_i = \text{yes} = \text{binomial parameter with}\: \pi_i
\end{cases}
$$

Then the model can be specified in ANOVA term.


\begin{equation}
\ln\frac{\pi_i}{1 - \pi_i} = \alpha + \beta_i, \: i = 1, 2, \ldots, I
\end{equation}


For redunduncies, we add a constraint. Among the three, we can choose anything.

\begin{equation}
\beta_1 = 0
\end{equation}

\begin{equation}
\beta_I = 0
\end{equation}

\begin{equation}
\sum_i \beta_i = 0
\end{equation}

### Two-way ANOVA representation

```{r}
aids
```

$$
\ln\frac{\pi_i}{1 - \pi_i} = \alpha + \beta_i^X + \beta_k^Z, \quad i = 1, \ldots, I,\: j = 1, \ldots, I
$$

constraint to $\beta_i^X$ and $\beta_i^Z$ among `(2)-(4)`. This model induces the relationship between $Y$ and $X$ given $Z$, i.e. conditional dependence.

## Indicator Variables

ANOVA-type model have presented various restrictions to allow parameters have non-negative degrees of freedom. Recall that in ANOVA, it might be important to construct _orthogonal design_. This leads to the following indicator variables coding for qualitative predictors.

### Dummy Coding

From `(2)` or `(3)`, we can implement so-called _dummy coding_. For example, `(3)` results in

| |$x_1$|$x_2$|$\cdots$|$x_{I-1}$|  
|:---:|:---:|:---:|:---:|:---:|  
|1|1|0|$\cdots$|0|  
|2|0|1|$\cdots$|0|  
|$\cdots$|$\cdots$|$\cdots$|$\cdots$|$\cdots$|  
|I-1|0|0|$\cdots$|1|  
|I|0|0|$\cdots$|0|  

```{r}
C(aids$azt %>% factor(),
  contr = contr.treatment, base = 2)
```

Here, dummy coding $\beta_2 = 0$ corresponds to

$$logit\pi = \beta_1 - \beta_2 = \beta_1$$

Thus, the estimate for reference category is the difference in logit (at a fixed level of $Z$).

On the other hand, when $\beta_1 = 0$ restriction is applied, the default `base = 1` can be used.

```{r}
C(aids$azt %>% factor(),
  contr = contr.treatment)
```

This can be interpreted as

$$logit\pi = \beta_1 - \beta_2 = -\beta_2$$

We might observe that the estimated coefficient will have reversed sign with the above $\beta_2 = 0$ coding.

### Effect Coding

From `(4)`, the last value can be coded as `-1`.

| |$x_1$|$x_2$|$\cdots$|$x_{I-1}$|  
|:---:|:---:|:---:|:---:|:---:|  
|1|1|0|$\cdots$|0|  
|2|0|1|$\cdots$|0|  
|$\cdots$|$\cdots$|$\cdots$|$\cdots$|$\cdots$|  
|I-1|0|0|$\cdots$|1|  
|I|-1|-1|$\cdots$|-1|  

```{r}
C(aids$azt %>% factor(),
  contr = contr.sum)
```

This corresponds to

$$logit\pi = \beta_1 - \beta_2 = 2\beta_1$$

The log odds ratio becomes twice of dummy coding induced by `(3)`. In terms of model parameter estimates, it would be the half of the dummy coding.

## Linear Logit Model for Contingency Tables

### Ordering categories

ANOVA-type model `(1)` is invariant to the factor-ordering.

\begin{equation}
logit(\pi_i) = \alpha + \beta x_i
\end{equation}

### Long data

To easily fit `glm`, we change the data to long format.

```{r}
(long_aids <-
  aids %>% 
  gather(yes, no, key = symptom, value = count) %>% 
  mutate(symptom = ifelse(symptom == "yes", 1, 0)) %>% 
  mutate_if(is.character, factor)) # to apply C() function
```


### Dummy Coding

By default, `base = 1` is used.

```{r}
C(long_aids$azt,
  contr = contr.treatment, base = 1)
C(long_aids$race,
  contr = contr.treatment, base = 1)
```

In an _alphabetical order_, `yes` and `white` each become the last category. We can manually set this with `level` option in `factor()` function. In `glm()`, there is an argument `contrasts = NULL`. This leads to `contr.treatment(base = 1)`.

$$AZT_{yes} = \begin{cases} 1 & \text{if yes} \\ 0 & \text{if no} \end{cases}$$

$$RACE_{white} = \begin{cases} 1 & \text{if white} \\ 0 & \text{if black} \end{cases}$$

```{r}
(dummy_last <-
  long_aids %>% 
  glm(symptom ~ azt + race, data = ., weights = count, family = binomial())) %>% 
  summary()
```

```{r}
dummy_last$contrasts
```

\begin{equation}
logit(\hat\pi) = -1.0736 - \underset{p-val = 0.0099}{0.7195} AZT_{yes} + \underset{p-val = 0.8476}{0.0555} RACE_{white}
\end{equation}

Controlling `racewhite`, we can say that `aztyes` significantly affects aids `symptom`.

### Dummy Coding with another `base`

By changing the dataset, we can freely implement the other qualitative coding. (Or we can specify the `contrasts` argument, e.g. `contrasts = c("contr.treatment", "contr.treatment")`.) We now try `contr.treatment(base = 2)`.

```{r}
C(long_aids$azt,
  contr = contr.treatment, base = 2)
C(long_aids$race,
  contr = contr.treatment, base = 2)
```

$$AZT_{no} = \begin{cases} 0 & \text{if yes} \\ 1 & \text{if no} \end{cases}$$

$$RACE_{black} = \begin{cases} 0 & \text{if white} \\ 1 & \text{if black} \end{cases}$$

```{r}
(dummy_first <-
  long_aids %>% 
  mutate_if(is.factor,
            funs(C(., contr = contr.treatment, base = 2))) %>% 
  glm(symptom ~ azt + race, data = ., weights = count, family = binomial())) %>% 
  summary()
```

```{r}
dummy_first$contrasts
```


\begin{equation}
logit(\hat\pi) = -1.0736 + \underset{p-val = 0.0099}{0.7195} AZT_{no} - \underset{p-val = 0.8476}{0.0555} RACE_{black}
\end{equation}

As mentioned, the absolute values of the estimates are exactly same. The only differences are _their signs_ by changing the definition of their variables.

### Effect Coding

In the same manner, `contr.sum` adjust $\sum\limits_i \beta_i = 0$ contraints.

```{r}
C(long_aids$azt,
  contr = contr.sum)
C(long_aids$race,
  contr = contr.sum)
```

$$AZT_{effect} = \begin{cases} 1 & \text{if yes} \\ -1 & \text{if no} \end{cases}$$

$$RACE_{effect} = \begin{cases} 1 & \text{if white} \\ -1 & \text{if black} \end{cases}$$

```{r}
(effect_sum <-
  long_aids %>% 
  mutate_if(is.factor,
            funs(C(., contr = contr.sum))) %>% 
  glm(symptom ~ azt + race, data = ., weights = count, family = binomial())) %>% 
  summary()
```

```{r}
effect_sum$contrasts
```

\begin{equation}
logit(\hat\pi) = -1.4056 + \underset{p-val = 0.01}{0.36} AZT_{effect} - \underset{p-val = 0.8476}{0.0277} RACE_{effect}
\end{equation}

```{r, echo=FALSE}
round(coef(effect_sum), 3) == round(coef(dummy_first) / 2, 3)
```

We can see the both $\beta_1$ and $\beta_2$ of `(8)` are half of the model `(7)`.

## Fitted values

Using the fitted logit model, we can estimate the expected frequency of contingency table. Given fitted model, we can estimate each conditional probabiltiy by

$$\hat\pi_{j \mid i} = P(Y_i = 1 \mid Z = j)$$

In `predict.glm()`, `type = "response"` option computes these probabilities. Also, the model object already has these values for the data set in `fitted.values`.

```{r}
dummy_last$fitted.values
dummy_first$fitted.values
effect_sum$fitted.values
```

Then the fitted number of successes(`yes`) becomes

$$\{ \hat\mu_{ij} = n_{i+}\hat\pi_{j \mid j} \}$$

```{r}
predict(dummy_last, type = "terms")
```





