```{r, message=FALSE}
library(tidyverse) # handling data
library(knitr) # formatting tables
library(kableExtra) # extra format -- multirow, xcolor, etc
```

```{r, include=FALSE}
# https://stackoverflow.com/questions/13848137/figure-captions-references-using-knitr-and-markdown-to-html
fig <- local({
    i <- 0
    ref <- list()
    list(
        cap = function(refName, text) {
            i <<- i + 1
            ref[[refName]] <<- stringr::str_c("Figure ", i) # in the text
            text # under the plot
        },
        ref = function(refName) {
            ref[[refName]]
        })
})
```

# Logit Model for Qualitative Predictors

```{r, message=FALSE}
aids <- read_table("data/AIDS.dat")
```

<!-- change to format = latex before compiling -->
<!-- Option clash for package xcolor: problem of knitr/kableExtra?? -->
<!-- works in Macbook pro 2013 late, but error in 2015 late: why?? -->

```{r, echo=FALSE}
aids %>% # https://haozhu233.github.io/kableExtra/awesome_table_in_html.html
  group_by(race) %>% 
  gather(yes, no, key = symptom, value = count, factor_key = TRUE) %>% # just to consider symptom columns at once
  mutate_if(is.numeric, function(x) {
    cell_spec(x, bold = TRUE,
              color = spec_color(x, begin = .3, end = .6),
              font_size = spec_font_size(-x, begin = 11))
  }) %>% 
  mutate(azt = cell_spec(
    azt, color = "white", bold = TRUE,
    background = spec_color(1:2, begin = .2, end = .7, option = "plasma", direction = 1)
  )) %>% 
  spread(symptom, count) %>% # return to the original set
  arrange(desc(race)) %>% # return to the original set
  kable(escape = FALSE, format = "html", row.names = FALSE, booktabs = TRUE,
        col.names = c("Race", "AZT Use", "Yes", "No"),
        caption = "Development of AIDS Symptoms by AZT Use and Race",
        align = "c") %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "HOLD_position", full_width = FALSE) %>% 
  add_header_above(header = c(" ", " ", "Symptoms" = 2)) %>% 
  collapse_rows(columns = 1)
```

Looking at the above table, direct usage of `azt` is likely to result in _slowing the development of AIDS symptoms_ (we can check this visually in `Table 1`). Our main interest is to analyze this relationship. To model this, we first define the binary response by

$$Y = \text{symptoms} = \begin{cases} \text{yes} = 1 \\ \text{no} = 0 \end{cases}$$

Denote that the (AZT)`azt` is also categorical predictor.

$$X = \text{AZT} = \begin{cases} \text{yes} \\ \text{no} \end{cases}$$

There is another factor `race` that has possibility to be covariate.

$$Z = \text{Race} = \begin{cases} \text{White} \\ \text{Black} \end{cases}$$

Based on our interest, we need to control the effect of this covariate.

## ANOVA-Type Representation of Factors

### One-way ANOVA representation

First consider a signle factor case, with $I$ categories(here, $I = 2$).

```{r}
aids %>% 
  select(-race)
```

For each row $i$ of the table, denote

$$
\begin{cases}
n_i = \text{yes} + \text{no} \\
y_i = \text{yes} = \text{binomial parameter with}\: \pi_i
\end{cases}
$$

Then the model can be specified in ANOVA term.

\begin{equation}
\ln\frac{\pi_i}{1 - \pi_i} = \alpha + \beta_i, \: i = 1, 2, \ldots, I
\end{equation}

For redunduncies, we add a constraint. Among the three, we can choose anything. We can set the frist term zero.

\begin{equation}
\beta_1 = 0
\end{equation}

Similarly, the last term can be set zero. Any other single $j$-term can be chosen.

\begin{equation}
\beta_I = 0
\end{equation}

By setting the whole sum as zero, we can guarantee the uniqueness.

\begin{equation}
\sum_i \beta_i = 0
\end{equation}

### Two-way ANOVA representation

```{r}
aids
```

$$
\ln\frac{\pi_i}{1 - \pi_i} = \alpha + \beta_i^X + \beta_k^Z, \quad i = 1, \ldots, I,\: j = 1, \ldots, I
$$

constraint to $\beta_i^X$ and $\beta_i^Z$ among `(2)-(4)`. This model induces the relationship between $Y$ and $X$ given $Z$, i.e. conditional dependence.

## Indicator Variables

ANOVA-type model have presented various restrictions to allow parameters have non-negative degrees of freedom. Recall that in ANOVA, it might be important to construct _orthogonal design_. This leads to the following indicator variables coding for qualitative predictors.

### Dummy Coding

From `(2)` or `(3)`, we can implement so-called _dummy coding_. For example, `(3)` results in

| |$x_1$|$x_2$|$\cdots$|$x_{I-1}$|  
|:---:|:---:|:---:|:---:|:---:|  
|1|1|0|$\cdots$|0|  
|2|0|1|$\cdots$|0|  
|$\cdots$|$\cdots$|$\cdots$|$\cdots$|$\cdots$|  
|I-1|0|0|$\cdots$|1|  
|I|0|0|$\cdots$|0|  

```{r}
C(aids$azt %>% factor(levels = c("yes", "no")),
  contr = contr.treatment, base = 2)
```

Here, dummy coding $\beta_2 = 0$ corresponds to

$$logit\pi = \beta_1 - \beta_2 = \beta_1$$

Thus, the estimate for reference category is the difference in logit (at a fixed level of $Z$).

On the other hand, when $\beta_1 = 0$ restriction is applied, the default `base = 1` can be used.

```{r}
C(aids$azt %>% factor(levels = c("yes", "no")),
  contr = contr.treatment)
```

This can be interpreted as

$$logit\pi = \beta_1 - \beta_2 = -\beta_2$$

We might observe that the estimated coefficient will have reversed sign with the above $\beta_2 = 0$ coding.

### Effect Coding

From `(4)`, the last value can be coded as `-1`.

| |$x_1$|$x_2$|$\cdots$|$x_{I-1}$|  
|:---:|:---:|:---:|:---:|:---:|  
|1|1|0|$\cdots$|0|  
|2|0|1|$\cdots$|0|  
|$\cdots$|$\cdots$|$\cdots$|$\cdots$|$\cdots$|  
|I-1|0|0|$\cdots$|1|  
|I|-1|-1|$\cdots$|-1|  

```{r}
C(aids$azt %>% factor(levels = c("yes", "no")),
  contr = contr.sum)
```

This corresponds to

$$logit\pi = \beta_1 - \beta_2 = 2\beta_1$$

The log odds ratio becomes twice of dummy coding induced by `(3)`. In terms of model parameter estimates, it would be the half of the dummy coding.

## Linear Logit Model for Contingency Tables

### Ordering categories

ANOVA-type model `(1)` is invariant to the factor-ordering.

\begin{equation}
logit(\pi_i) = \alpha + \beta x_i
\end{equation}

### Long data

To easily fit `glm`, we change the data to long format. We can handle `symptom` in two way. Using `factor` or dichotomous `1-0` numeric. `CRAN documentation` for `binomial()` link function gives those approach.

> As a numerical vector with values between 0 and 1, interpreted as the proportion of successful cases (with the total number of cases given by the weights).

For this, we set development of AIDS as $Y = 1$, otherwise $Y = 0$.

```{r}
aids %>% 
  gather(yes, no, key = symptom, value = count) %>% 
  mutate(symptom = ifelse(symptom == "yes", 1, 0)) %>% 
  mutate_if(is.character, factor) # to apply C() function
```

> As a factor: ‘success’ is interpreted as the factor not having the first level (and hence usually of having the second level).

This means that if we put `no` followed by `yes`, `glm()` will fit $P(AIDS = yes)$.

```{r}
(long_aids <-
  aids %>% 
  gather(no, yes, key = symptom, value = count, factor_key = TRUE) %>% # yes after no
  mutate_if(is.character, function(x) {
    factor(x, levels = unique(x)) # levels = same order as in data
  }))
```

### Dummy Coding induce by `(2)`

By default, `base = 1` is used.

```{r}
C(long_aids$azt,
  contr = contr.treatment, base = 1)
C(long_aids$race,
  contr = contr.treatment, base = 1)
```

`factor()` choose its level by an _alphabetical order_. Here we have set the levels manually with `levels` option in the function, which are the same as the order of the data. In `glm()`, there is an argument `contrasts = NULL`. This leads to `contr.treatment(base = 1)`.

$$AZT_{no} = \begin{cases} 0 & \text{if yes} \\ 1 & \text{if no} \end{cases}$$

$$RACE_{black} = \begin{cases} 0 & \text{if white} \\ 1 & \text{if black} \end{cases}$$

```{r}
(dummy_first <-
  long_aids %>% 
  glm(symptom ~ azt + race, data = ., weights = count, family = binomial())) %>% 
  summary()
```

```{r}
dummy_first$contrasts
```

```{r, include=FALSE}
dummy1 <- broom::tidy(dummy_first)
```


\begin{equation}
logit(\hat\pi) = `r dummy1$estimate[1]`  \underset{p-val = `r dummy1$p.value[2]`}{`r dummy1$estimate[2]`} AZT_{no} \underset{p-val = `r dummy1$p.value[3]`}{`r dummy1$estimate[3]`} RACE_{black}
\end{equation}

Controlling `raceblack`, we can say that `aztno` significantly affects aids `symptom`.

### Dummy Coding induced by `(3)`

By changing the dataset, we can freely implement the other qualitative coding. We now try `contr.treatment(base = 2)`, i.e. setting the last term zero.

```{r}
C(long_aids$azt,
  contr = contr.treatment, base = 2)
C(long_aids$race,
  contr = contr.treatment, base = 2)
```

$$AZT_{yes} = \begin{cases} 1 & \text{if yes} \\ 0 & \text{if no} \end{cases}$$

$$RACE_{white} = \begin{cases} 1 & \text{if white} \\ 0 & \text{if black} \end{cases}$$

```{r}
make_dummy <- function(x, contr, ...) { # for coefficients' names
  new_x <- C(x, contr = contr, ...)
  attr_x <- attributes(new_x)$contrasts
  colnames(attributes(new_x)$contrasts) <- rownames(attr_x)[attr_x > 0]
  new_x
}
```

Applying `C(contr = contr.treatment, base = 2)`, we get

```{r}
(dummy_last <-
  long_aids %>% 
  mutate_at(.vars = vars(azt, race),
            .funs = funs(make_dummy(., contr = contr.treatment, base = 2))) %>% 
  glm(symptom ~ azt + race, data = ., weights = count, family = binomial())) %>% 
  summary()
```

```{r}
dummy_last$contrasts
```


\begin{equation}
logit(\hat\pi) = -1.0736 - \underset{p-val = 0.0099}{0.7195} AZT_{yes} + \underset{p-val = 0.8476}{0.0555} RACE_{white}
\end{equation}

As mentioned, the absolute values of the estimates are exactly same. The only differences are _their signs_ by changing the definition of their variables.

### Effect Coding

In the same manner, `contr.sum` adjust $\sum\limits_i \beta_i = 0$ contraints. Or we can specify the `contrasts` argument, e.g. `contrasts = list(azt = "contr.sum", race = "contr.sum")`.

```{r}
C(long_aids$azt,
  contr = contr.sum)
C(long_aids$race,
  contr = contr.sum)
```

$$AZT_1 = \begin{cases} 1 & \text{if yes} \\ -1 & \text{if no} \end{cases}$$

$$RACE_1 = \begin{cases} 1 & \text{if white} \\ -1 & \text{if black} \end{cases}$$

```{r}
(effect_sum <-
  long_aids %>% 
  glm(symptom ~ azt + race, data = ., weights = count, family = binomial(),
      contrasts = list(azt = "contr.sum", race = "contr.sum"))) %>% 
  summary()
```


```{r}
effect_sum$contrasts
```

\begin{equation}
logit(\hat\pi) = -1.4056 - \underset{p-val = 0.01}{0.36} AZT_{effect} - \underset{p-val = 0.8476}{0.0277} RACE_{effect}
\end{equation}

```{r, echo=FALSE}
round(coef(effect_sum), 3) == round(coef(dummy_last) / 2, 3)
```

We can see the both $\beta_1$ and $\beta_2$ of `(8)` are half of the model `(7)`.

## Confidence Intervals

<!-- MASS::confint versus confint.default -->

`confint()` gives CI computed by profile likelihood, if `MASS` package is installed.

```{r, message=FALSE}
lapply(list(First_zero = dummy_first, Last_zero = dummy_last, Sum_zero = effect_sum), confint)
```

or `confint.default()` gives CI computed by the standard error.

```{r}
lapply(list(First_zero = dummy_first, Last_zero = dummy_last, Sum_zero = effect_sum), confint.default)
```

## Fitted values

Using the fitted logit model, we can estimate the expected frequency of contingency table. Before that, estimate each success conditional probabiltiy, i.e. probability of AIDS development by

$$\hat\pi_{j \mid i} = P(Y_i = 1 \mid Z = j) = \frac{\exp(\hat\alpha + \hat\beta_1X + \hat\beta_2Z)}{1 + \exp(\hat\alpha + \hat\beta_1X + \hat\beta_2Z)}$$

`predict()` for `glm` object gives various values with `type` option.

1. By default, `type = "link"` gives the linear fit on link scale: this is same as `fit$linear.predictors`
2.  `type = "response"`: on the scale of the response variable, here gives probabilities on logit we want = `fit$fitted.values`
3. `type = "terms"`: on the scale of linear predictor scale


```{r}
predict(dummy_last, type = "response")
dummy_last$fitted.values
```

```{r}
long_aids %>% 
  mutate(pred_dummy1 = predict(dummy_last,
                               newdata = data_frame(race = race, azt = azt),
                               type = "response"),
         pred_dummy2 = predict(dummy_last,
                               newdata = data_frame(race = race, azt = azt),
                               type = "response"),
         pred_effect = predict(dummy_last,
                               newdata = data_frame(race = race, azt = azt),
                               type = "response")) %>% 
  spread(symptom, count) %>% # return to contingency table format
  select(race, azt, yes, no,
         pred_dummy1, pred_dummy2, pred_effect)
```

In fact, coding does not affect goodness-of-fit.

```{r probaids, fig.cap=fig$cap("probaids", "Conditional probability of developing AIDS symptoms")}
(pred_prob <-
  long_aids %>% 
  mutate(pred_prob = predict(dummy_last,
                               newdata = data_frame(race = race, azt = azt),
                               type = "response")
         ) %>% 
  spread(symptom, count) %>% # return to contingency table format
  select(race, azt, yes, no, pred_prob)) %>% 
  gather(yes, no, key = symptom, value = count) %>% # to plot
  ggplot() +
  aes(x = race, y = pred_prob, group = azt) +
  geom_line(aes(colour = azt)) +
  labs(
    x = "Race",
    y = expression(pi),
    parse = TRUE
  )
```

Obviously, less probability is likely occur when the patient dose AZT directly. In `r fig$ref("probaids")`, the probability differs between AZT usage, while not between race. Now we can estimate the number of successes by

$$\{ \hat\mu_{ij} = n_{i+}\hat\pi_{j \mid i} \}$$

```{r}
pred_prob %>% 
  mutate(yes_fit = (yes + no) * pred_prob,
         no_fit = (yes + no) * (1 - pred_prob)) %>% 
  arrange(desc(race), desc(azt)) %>% 
  select(race, azt, yes, yes_fit, no, no_fit, pred_prob)
```

It can be estimated as

```{r}
pred_prob %>% 
  mutate(yes_fit = (yes + no) * pred_prob %>% round(digits = 1),
         no_fit = (yes + no) * (1 - pred_prob) %>% round(digits = 1)) %>% 
  unite("Yes", starts_with("yes"), sep = " vs ") %>% 
  unite("No", starts_with("no"), sep = " vs ")
```

In sum,

```{r, echo=FALSE}
pred_prob %>% 
  mutate(yes_fit = round((yes + no) * pred_prob, digits = 1),
         no_fit = round((yes + no) * (1 - pred_prob), digits = 1),
         pred_no = round(1 - pred_prob, digits = 3),
         pred_prob = round(pred_prob, digits = 3)) %>% 
  select(race, azt, pred_prob, pred_no, yes_fit, no_fit) %>% 
  gather(yes_fit, no_fit, key = fit, value = count) %>%
  gather(pred_prob, pred_no, key = prob, value = pred) %>%
  mutate_if(is.numeric, function(x) {
    cell_spec(x, bold = TRUE,
              color = spec_color(-x, end = .9),
              font_size = spec_font_size(x, begin = 10))
  }) %>% 
  group_by(race) %>% 
  mutate(azt = cell_spec(
    azt, color = "white", bold = TRUE,
    background = spec_color(1:2, begin = .2, end = .7, option = "plasma", direction = 1)
  )) %>% 
  ungroup(race) %>% 
  spread(prob, pred) %>% 
  spread(fit, count) %>% 
  select(race, azt, pred_prob, pred_no, yes_fit, no_fit) %>% 
  kable(escape = FALSE, format = "latex", row.names = FALSE, booktabs = TRUE,
        col.names = c("Race", "AZT Use", "Yes", "No", "Yes", "No"),
        caption = "Estimated probability and Fitted number",
        align = "c") %>% 
  kable_styling(bootstrap_options = "striped", latex_options = "HOLD_position", full_width = FALSE) %>% 
  add_header_above(header = c(" ", " ", "Fitted Probability" = 2, "Fitted number" = 2))
```

<!-- round probabilities -->

## Appendix {-}

We can also use contingency table form for `glm()`

```{r}
aids %>% 
  mutate(Sum = yes + no) %>% 
  glm(yes/Sum ~ azt + race, data = ., weights = Sum, family = binomial())
```








