```{r, message=FALSE}
library(tidyverse)
```

# Loglinear Models for Contingency Tables

## Loglinear models for Two-way tables

> Both variables are response variables.

### Association between responses

Consider $I \times J$ contingency table $\{ n_{ij} \}$. Then

$$n_{ij} \sim Poisson(\mu_{ij})$$


## Loglinear models for Three-way tables

## Alcohol, cigarette, and marijuana use {-}

```{r, message=FALSE}
(substance <-
  read_delim("data/Substance_use.dat", delim = " ") %>% 
  mutate(alcohol = str_trim(alcohol))) # due to messy data file
```

To fit loglinear model, long data format like this is easy.

```{r}
substance <-
  substance %>% 
  mutate_if(is.character, factor) %>% 
  mutate_if(is.factor, fct_rev)
```

Below is _mutual independence model_ `(A, C, M)`.

```{r}
(subs_log <-
  substance %>% 
  glm(count ~ alcohol + cigarettes + marijuana, data = ., family = poisson())) %>% 
  summary()
```

### Fitted values

```{r}
fit_loglin <- function(x, ...) {
  mod_name <-
    as.character(x[[1]]$call)[2] %>% 
    str_extract(pattern = "(?<=~).*") %>% 
    str_trim()
  x[[1]]$model %>% 
    bind_cols(predict(x[[1]], newdata = ., type = "response", ...) %>% tbl_df()) %>% 
    rename_at(.vars = vars(value), .funs = funs(return(mod_name)))
}
#--------------------------------
(subs_fit <-
  subs_hierarchy %>% 
  map(fit_loglin) %>% 
  plyr::join_all(by = c("count", "alcohol", "cigarettes", "marijuana"))) %>% 
  pander::pander()
```

### Chi-square Goodness-of-fit tests

$$G^2 = 2\sum n_{ijk}\ln\frac{n_{ijk}}{\hat\mu_{ijk}}$$

$$X^2 = \sum\frac{(n_{ijk} - \hat\mu_{ijk})^2}{\hat\mu_{ijk}}$$

with

$$\text{residual df} = \text{the number of cell counts} - \text{the number of non-redundant parameters}$$

```{r}
subs_hierarchy <-
  substance %>% 
  do(
    indep = glm(count ~ alcohol + cigarettes + marijuana, data = ., family = poisson()),
    ac_m = glm(count ~ alcohol + cigarettes + marijuana + alcohol:cigarettes, 
               data = ., family = poisson()),
    amcm = glm(count ~ alcohol + cigarettes + marijuana + alcohol:marijuana + cigarettes:marijuana, 
               data = ., family = poisson()),
    acamcm = glm(count ~ alcohol + cigarettes + marijuana + alcohol:cigarettes + alcohol:marijuana + cigarettes:marijuana, 
               data = ., family = poisson()),
    acm = glm(count ~ alcohol * cigarettes * marijuana, 
                 data = ., family = poisson())
  )
```


```{r, message=FALSE, warning=FALSE}
good_loglin <- function(x, test = "LRT", ...) {
  mod_name <-
    as.character(x[[1]]$call)[2] %>% 
    str_extract(pattern = "(?<=~).*") %>% 
    str_trim()
  broom::tidy(anova(x[[1]], test = test, ...)) %>% 
    slice(n()) %>% 
    add_column(model = mod_name, .before = 1) %>% 
    select(-term)
}
#-----------------------------------------------
(subs_good <-
  subs_hierarchy %>% 
  map(good_loglin, test = "LRT") %>% 
  bind_rows()) %>% 
  pander::pander()
```

From above $G^2$, we compare reduced model to complex model

$$G^2(M_0 \mid M_1) = G^2(M_0) - G^2(M_1) \approx \chi^2\Big(df = df(M_0) - df(M_1)\Big)$$

```{r}
subs_good %>% 
  select(-Deviance, -p.value) %>% 
  rename(alternative = model) %>% 
  mutate(goodness = c(Resid..Dev[1], -diff(Resid..Dev)),
         df_good = c(Resid..Df[1], -diff(Resid..Df))) %>% 
  mutate(p_value = pchisq(goodness, df = df_good, lower.tail = FALSE)) %>% 
  pander::pander()
```

1. saturated model `(ACM)`: cannot reject $M_0$, so we choose next model
2. three factor interaction `(AC,AM,CM)`: reject $M_0$

Thus, we use model _(AC, AM, CM)_.










