```{r, include=FALSE}
library(tidyverse)
```

# Generalized Linear Models for Counts and Rates

It is natural to assume the count data

$$Y \sim Poisson(\mu)$$

## Poisson Loglinear Models

As mentioned, we build a GLM with

1. The _random component_: $Y \sim Poisson(\mu)$
2. The _systematic component_: $\alpha + \beta_1 x_1 + \cdots + \beta_p x_p$
3. The _link function_: log-link $\ln(\mu)$ which is canonical link for a Poisson GLM

$$\ln\mu(\mathbf{x}) = \alpha + \beta_1x_1 + \cdots + \beta_px_p$$

For an interpretation perspective,

$$\mu(\mathbf{x}) = \exp(\alpha + \beta_1x_1 + \cdots + \beta_px_p) = e^\alpha(e^{\beta_1})^{x_1}\cdots(e^{\beta_p})^{x_p}$$

i.e. $\mu$ is multiplied by $e^{\beta_j}$ as $x_j$ increases by 1-unit. $x_j$ has a _multiplicative impact_ of $e^{\beta_j}$ on $\mu$.


## Horseshoe crab mating data

```{r, message=FALSE}
(crab <- read_table("data/Crabs.dat"))
```

```{r}
crab %>%
  ggplot() +
  aes(x = width, y = sat) +
  geom_hex() +
  labs(
    x = "Carapace width",
    y = "Number of satellites"
  )
```

- Large variability is observed.
- Outlier: analysis with this observation and without it

For clarification,

```{r}
crab %>%
  group_by(width = cut_interval(width, n = 8))
```













