```{r, include=FALSE}
library(tidyverse)
```

# Generalized Linear Models for Counts and Rates

It is natural to assume the count data

$$Y \sim Poisson(\mu)$$

## Poisson Loglinear Models

As mentioned, we build a GLM with

1. The _random component_: $Y \sim Poisson(\mu)$
2. The _systematic component_: $\alpha + \beta_1 x_1 + \cdots + \beta_p x_p$
3. The _link function_: log-link $\ln(\mu)$ which is canonical link for a Poisson GLM

$$\ln\mu(\mathbf{x}) = \alpha + \beta_1x_1 + \cdots + \beta_px_p$$

For an interpretation perspective,

$$\mu(\mathbf{x}) = \exp(\alpha + \beta_1x_1 + \cdots + \beta_px_p) = e^\alpha(e^{\beta_1})^{x_1}\cdots(e^{\beta_p})^{x_p}$$

i.e. $\mu$ is multiplied by $e^{\beta_j}$ as $x_j$ increases by 1-unit. $x_j$ has a _multiplicative impact_ of $e^{\beta_j}$ on $\mu$.


## Horseshoe crab mating data

```{r, message=FALSE}
(crab <- read_table("data/Crabs.dat"))
```

```{r}
crab %>%
  ggplot() +
  aes(x = width, y = sat) +
  geom_hex() +
  labs(
    x = "Carapace width",
    y = "Number of satellites"
  )
```

- Large variability is observed.
- Outlier: analysis with this observation and without it

For clarification,

```{r}
(width_interval <-
  crab %>%
  group_by(width = cut(width,
                       breaks = c(0, seq(23.25, 29.25, by = 1), Inf),
                       ordered_result = TRUE)) %>%
  summarise(cases = n(), S = sum(sat), Mean = mean(sat), Variance = var(sat)))
```

Looking at the table, we can see the _nonlinear relationship_ between sateliite counts and width.



```{r}
crab %>%
  group_by(width = cut(width,
                       breaks = c(0, seq(23.25, 29.25, by = 1), Inf),
                       ordered_result = TRUE)) %>%
  ggplot() +
  aes(x = width, y = sat, fill = width) +
  geom_boxplot(show.legend = FALSE)
```

### Logistic regression model

$$Y = \begin{cases} 1 & \text{if satellite}\: > 0 \\ 0 & \text{if satellite}\: = 0 \end{cases}$$

```{r}
crab %>%
  select(y, width)
```


```{r}
logistic_fit <-
  crab %>%
  glm(y ~ width, data = ., family = binomial(link = "logit"))
summary(logistic_fit)
```


```{r}
anova(logistic_fit, test = "LRT")
```



### Poisson regression

```{r}
crab %>%
  select(sat, width)
```


```{r}
pois_fit <-
 crab %>%
  glm(sat ~ width, data = ., family = poisson(link = "log"))
summary(pois_fit)
```


```{r}
anova(pois_fit, test = "LRT")
```


### Overdispersion

```{r}
width_interval
```

`Variance` is much larger than `Mean`.


### Negative Binomial GLMs




















